Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.tex
partial comments on section 2 rewrite and other clean-up

\textcolor{violet}{I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful. Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.}

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
and visualizations from universities, many branches and levels of
government, and news organizations are an important first step of the
process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021} .

With the many visualization and exploratory tools, models and modeling
paradigms, and reviews and comparisons in the literature and through the
MIDAS (Models of Infectious Disease Agent Study) network
\citep{midasnetwork2021}, this field has a lot of devices to aid an
individual practitioner decide the correct approach. For
example,\proglang{R} packages such as \pkg{surveillance},
\pkg{EpiModel}, and \pkg{pomp} have all made significant steps in
standardizing the flow of the data analysis pipeline for epidemic
modeling through digitizing data sets, making accessible statistical
models, and providing a plethora of educational material for both coding
novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.

At the same time, analysis packages often only address a specific
portion of the analysis pipeline, for instance focusing on certain types
of models. Modeling tools, which usually require learning
package-specific syntax, often don't provide easy ways to compare and
assess their models on new data. Moreover, exploring and modeling
epidemics require transforming and \textit{tidying} data in different
ways. To fill these gaps, we present our \proglang{R} package
\pkg{EpiCompare}. Our package's primary focus is to aid and advance
research in the area of comparison and assessment of epidemic and
epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the
data analysis pipeline of infectious diseases as 1) data pre-processing,
2) exploratory data analysis (EDA), 3) modeling and simulating, 4)
post-processing, and 5) comparison and assessment; where each previous
part of the pipeline influences the next. \pkg{EpiCompare} provides
tools to aids practitioners in all areas of this pipeline.

@@

\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way. Epidemics, despite by definition being a
process that evolves over time, often need to be compared in a way not
constrained to initial times or time scales to understand the processes
at play.
\textcolor{violet}{Moreover, many current comparison tools for state-space models (e.g. SIR models) highlight the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion. This approach may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.}
Tools in \pkg{EpiCompare} give the user the ability to extend their
toolkit to evaluate epidemics within a time-invariant lens. The goal of
\pkg{EpiCompare} is not to supplant existing infectious disease modeling
tools and software but, rather, is a concerted effort to create standard
and fair comparisons among models developed for disease outbreaks and
outbreak data.

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a thorough demonstrating of the tools
through a detailed example of a full data analysis pipeline.

@@

\textcolor{violet}{the vast majority of this section has been changed / needs more updates.}

@@

\textcolor{orange}{I again changed this section a lot, but hopefully it's more in line with what you are thinking.  The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot; and 4) talk about models beyond SIR.}

@@

\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by
(1) taking a global, not marginal view of how epidemics move through
\sout{time} \textcolor{orange}{populations} and (2) by treating full
epidemics as
filaments\sout{ not functions (breaking away from a time-centric view)}
and not points produced by functions of time. The following section aims
to highlight the strengths of \textit{time-invariance} and define the
mathematical foundations that \pkg{EpiCompare}'s tools stand upon.

\textcolor{orange}{Mathematically}, epidemics are complex objects. They
are hard to assess and compare to one another due to the differences in
the diseases, the location where the outbreak occurs, how the
\textcolor{orange}{a?}ffected population reacts, and the timefeatures
(including start of the epidemic, speed of infection and more).
Time-invariant analysis \sout{attempts to make}
\textcolor{orange}{makes} different epidemics easier to compare by
removing many time dependent aspects of an epidemic. Instead
\textcolor{orange}{it} focuses on the overall \sout{"life"}
\textcolor{orange}{trajectory} of an epidemic, emphasizing the number of
lives affected.
\textcolor{orange}{really like the lives affected part; I think that's a powerful statement}

@@

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we illustrate how our
\textcolor{violet}{\sout{package fits} package's functions fit} into the
data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions are aimed to be as user-friendly as possible. We
also focus on providing the user ``tidyverse'' style functions, that
encourage piping and also follow clear verb naming schemes
\citep{Wickham2019}. Although users can typically incorporate
\pkg{EpiCompare} into any step in the data analysis process, there are
two primary points of entry. The first point of entry is the very
beginning with pre-processing and visualizing raw data, and the second
point of entry is after modeling and simulation. Figure
\ref{fig:pipeline2} captures these different paths, and we will
highlight both approaches and how to leverage \pkg{EpiCompare} in the
subsections below.

@@

The first step of most data analysis is cleaning the data
\textcolor{violet}{so it can \sout{to}} be explored. There are multiple
ways to collect epidemiological data. Sometimes individual records
\textcolor{violet}{are collected}, with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and sub-population
information\textcolor{violet}{.\sout{will be collected, whereas}}
\textcolor{violet}{O}ther data collections focus on aggregate counts of
individuals in each epidemic state. In fact, usually only the number of
new infections at each time step (e.g.~weekly case counts) is observed.
Compartment totals
\textcolor{violet}{(amounts of individuals in each state)} are then
imputed from those case counts along with other information about the
disease and the population of interest.
\textcolor{violet}{In \pkg{EpiCompare}, w}e focus on understanding the
overall impact of an outbreak at the aggregate/population level,
\textcolor{violet}{which allows for examination of overall trends of an epidemic more easily}.

\textcolor{violet}{[Ben says: the following rewrite was done because the original text seemed to repeat the same thing 2x, once in text and the second time with the function }\texttt{agents\_to\_aggregate}\textcolor{violet}{. The following two paragraphs try to merge the introduction of the function and it's capabilities into one story.]}

\textcolor{violet}{[Ben says: this whole paragraph is new]} In order to
help the practitioner examine epidemics from an aggregate/ population
lens, we provide a function called \texttt{agents\_to\_aggregate}. This
function transforms data about individual/agents' initial entry into
each state (e.g.~start of infection, start of recovery, etc.) to an
aggregate view of how many individuals were in a state at a given time.
There are often situations where grouping agents into subpopulations
(e.g.~subpopulations defined by age or sex) can highlight different
aggregate level trends. For example, research by
\citet{rvachev1985,anderson1992,worby2015} develop state-based models
that account for subpopulations. In \pkg{EpiCompare}, we facilitate
subpopulation analysis by combining \pkg{dplyr}::\texttt{group\_by} and
\texttt{agent\_to\_aggregate} to provide aggregation at a group level.

\textcolor{violet}{[Ben says: this whole paragraph is new]} The
\texttt{agents\_to\_aggregate} function is flexible and can deal with a
wide range of information about each individual. It can, theoretically,
account for infinitely many states. This functionality allows the
practitioner to aggregate information relative to the standard states
(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as add
states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'') (CITE).
Additionally, \texttt{agents\_to\_aggregate} also permits indicators for
death/exit and birth/entry dates. Overall, the function
\texttt{agents\_to\_aggregate()} is a powerful tool for pre-processing
data.

\textcolor{violet}{\sout{In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g. start of infection, start of recovery, etc). This transformation between \textit{agent} information to \textit{aggregate} information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations). Our tools allow the user to easily group agents and define new subpopulations to explore. This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g. age, sex) and many state based models provide for subpopulation-based states in their analysis}}
\citep{rvachev1985,anderson1992,worby2015}.

\textbackslash ben\{\sout{We provide a "tidyverse"-styled function, }\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ to transform agent information into aggregate state information.  As a "tidy" function, our function }}\texttt{agents\_to\_aggregate()}
\textcolor{violet}{\sout{allows the user to identify and analyze subpopulations by first executing a }}\texttt{group\_by()}
\textbackslash ben\{\sout{{ from \pkg{dplyr}. The function }}\texttt{agents\_to\_aggregate}\textcolor{violet}{\sout{is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates. Our function,}}
\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ does not require agents to pass through every state listed and agents are allowed to start in any state. This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g. hours or minutes). The function }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ returns the total number of individuals in each state at each time point.  We view }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.}}

@@

\textcolor{violet}{[Ben says: I'm went back and thought that the version before this version was clearer. Wanna take a look and discuss? Side by side: [github](https://github.com/skgallagher/EpiCompare/commit/7dbb4996d1d3169d3e3a426bc9820c417ea87beb)]}

\textcolor{violet}{[Ben says: this is actually an older version:]} With
raw data, ``getting to know'' our data currently means figuring out good
combinations of visualizations, numerical summaries and subsets. An
expert coder has many ways to successfully explore the data in an
aggregate lens using \texttt{agents\_to\_aggregate}. \pkg{EpiCompare}
also includes tools to rapidly explore data that has three epidemic
states. Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our
\texttt{geom\_aggregate} provides a rapid way to explore different
subpopulations' experience of an epidemic by combining the ideas behind
\texttt{agents\_to\_aggregate} for three-state models to examine
subpopulation trajectories in a 3d simplex space
\citep{Wickham2016, Hamilton2018}. \textcolor{orange}{come back?}
Visualization tools for three-state models were developed because (1)
SIR models are some of the most common and basic epidemic state-based
models and (2) our simplex representation of these epidemics emphasizes
a ``time-invarance'' representation of the data (for a refresher see
Section \ref{sec:time-invariant}).
\textcolor{orange}{make sure SIR is defined before}

\textcolor{violet}{[Ben says: The 2 things I really like about the earlier (now current) version is (1) the introduction seems to make more sense in getting at what EDA really is, (2) the way it disucsses SIR models only 1 potential model instead of putting it as *the* model.]}

\textcolor{violet}{\sout{With raw data, exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings. An expert coder can start from}}
\texttt{agents\_to\_aggregate()}
\textcolor{violet}{\sout{to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model). Our}}
\texttt{geom\_aggregate()}
\textcolor{violet}{\sout{provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens. It combines the ideas behind}}
\texttt{agents\_to\_aggregate()}
\textbackslash ben\{\sout{with \pkg{ggplot2}'s} \texttt{geom\_path()}
\textcolor{violet}{\sout{and \pkg{ggtern}'s}} \texttt{coord\_tern()}
\textcolor{violet}{\sout{to visualize any number of groups epidemic trajectory in 3d simplex}}
\citep{Wickham2016, Hamilton2018}.\textcolor{violet}{\sout{We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).}}

>>>>

\textcolor{violet}{[Ben says: I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful. Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.]}

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
and visualizations from universities, many branches and levels of
government, and news organizations are an important first step of the
process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.]}

With the many visualization and exploratory tools, models and modeling
paradigms, and reviews and comparisons in the literature and through the
MIDAS (Models of Infectious Disease Agent Study) network
\citep{midasnetwork2021}, this field has a lot of devices to aid an
individual practitioner decide the correct approach. For example,
\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
\pkg{pomp} have all made significant steps in standardizing the flow of
the data analysis pipeline for epidemic modeling through digitizing data
sets, making accessible statistical models, and providing a plethora of
educational material for both coding novices and experts alike
\citep{surveillance2017,Jenness2018,King2016}.

At the same time, analysis packages often only address a specific
portion of the analysis
pipeline\textcolor{violet}{\sout{, for instance focusing on certain types of models.}}
\textcolor{violet}{These m}odeling
tools\textcolor{violet}{\sout{, which}} usually require learning
package-specific syntax\textcolor{violet}{\sout{,}} and often don't
provide easy ways to compare and assess their models on new data.
Moreover, exploring\textcolor{violet}{, \sout{and}} modeling
\textcolor{violet}{and comparing} epidemics require transforming and
\textit{tidying} data in different ways. To fill these gaps, we present
our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
is to aid and advance research in the area of comparison and assessment
of epidemic and epidemiological models. In Figure \ref{fig:pipeline}, we
illustrate the data analysis pipeline of infectious diseases as 1) data
pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
simulating, 4) post-processing, and 5) comparison and assessment; where
each previous part of the pipeline influences the next. \pkg{EpiCompare}
provides tools to aids practitioners in all areas of this pipeline.

@@

\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way. Epidemics, despite by definition being a
process that evolves over time, often need to be compared in a way not
constrained to initial times or time scales to understand the processes
at play.
\textcolor{violet}{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion. \sout{Moreover, m} M}any
current \textcolor{violet}{time-dependent} comparison tools for
state-space models (e.g.~SIR models)
\textcolor{violet}{\sout{highlight} examine} the proportion of
individuals in each state (at a given time) in a piece-wise / marginal
fashion. \textcolor{violet}{These \sout{This}}
approach\textcolor{violet}{es} may reduce the amount of connections that
can be seen, similar to projections of a multidimensional distribution
onto a single axis at a time. Tools in \pkg{EpiCompare} give the user
the ability to extend their toolkit to evaluate epidemics within a
time-invariant lens. The goal of \pkg{EpiCompare} is not to supplant
existing infectious disease modeling tools and software but, rather, is
a concerted effort to create standard and fair comparisons among models
developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a \textcolor{violet}{\sout{thorough}}
demonstrating of the tools through a detailed example of a full data
analysis pipeline.

@@

@@

\textcolor{orange}{I again changed this section a lot, but hopefully it's more in line with what you are thinking.  The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot\footnote{\textcolor{violet}{[Ben says: defend how this fits into the narrative?]}} and 4) talk }\textcolor{violet}{Time invariance in}
\sout{about}\textcolor{orange}{ models beyond SIR.}

@@

\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by
(1) taking a global, not marginal view of how epidemics move through
populations and (2) by treating full epidemics as filaments\footnote{\textcolor{violet}{[Ben says: Do we need a footnote explaining what this is more clearly?]}}
and not points produced by functions of time. The following section aims
to highlight the strengths of \textit{time-invariance} and define the
mathematical foundations that \pkg{EpiCompare}'s tools stand upon.

Mathematically, epidemics are complex objects. They are hard to assess
and compare to one another due to the differences in the diseases, the
location where the outbreak occurs, how the affected population reacts,
and the time \textcolor{violet}{related} features (including start of
the epidemic, speed of infection and more). Time-invariant analysis make
different epidemics easier to compare by removing many time dependent
aspects of an epidemic. Instead it focuses on the overall trajectory of
an epidemic, emphasizing the number of lives affected.
\textcolor{orange}{[Shannon says: really like the lives affected part; I think that's a powerful statement.]}

@@

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we illustrate how our package's functions fit into
the data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions are aimed to be as user-friendly as possible. We
also focus on providing the user ``tidyverse'' style functions, that
encourage piping and also follow clear verb naming schemes
\citep{Wickham2019}. Although users can typically incorporate
\pkg{EpiCompare} into any step in the data analysis process, there are
two primary points of entry. The first point of entry is the very
beginning with pre-processing and visualizing raw data, and the second
point of entry is after modeling and simulation. Figure
\ref{fig:pipeline2} captures these different paths, and we will
highlight\footnote{\textcolor{violet}{[Ben says: we need to make sure we actually do highlight ]}}
both approaches and how to leverage \pkg{EpiCompare} in the subsections
below.

@@

The first step of most data analysis is cleaning the data so it can be
explored. There are multiple ways to collect epidemiological data.
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact, usually only the number of
new infections at each time step (e.g.~weekly case counts) is observed.
Compartment totals (amounts of individuals in each state) are then
imputed from those case counts along with other information about the
disease and the population of interest. In \pkg{EpiCompare}, we focus on
understanding the overall impact of an outbreak at the
aggregate/population level, which allows for examination of overall
trends of an epidemic more easily.

In order to help the practitioner examine epidemics from an aggregate/
population lens, we provide a function called
\texttt{agents\_to\_aggregate}. This function transforms data about
individual/agents' initial entry into each state (e.g.~start of
infection, start of recovery, etc.) to an aggregate view of how many
individuals were in a state at a given time. There are often situations
where grouping agents into subpopulations (e.g.~subpopulations defined
by age or sex) can highlight different aggregate level trends. For
example, research by \citet{rvachev1985,anderson1992,worby2015} develop
state-based models that account for subpopulations. In \pkg{EpiCompare},
we facilitate subpopulation analysis by combining
\pkg{dplyr}::\texttt{group\_by} and \texttt{agent\_to\_aggregate} to
provide aggregation at a group level.

The \texttt{agents\_to\_aggregate} function is flexible and can deal
with a wide range of information about each individual. It can,
theoretically, account for infinitely many states. This functionality
allows the practitioner to aggregate information relative to the
standard states (e.g.~``Susceptible'', ``Infectious'', and
``Recovered'') as well as add states (e.g.~``Exposed'', ``iMmune'',
``Hospitalized'')
(CITE\footnote{\textcolor{violet}{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}}).
Additionally, \texttt{agents\_to\_aggregate} also permits indicators for
death/exit and birth/entry dates. Overall, the function
\texttt{agents\_to\_aggregate()} is a powerful tool for pre-processing
data.

@@

With raw data, ``getting to know'' our data currently means figuring out
good combinations of visualizations, numerical summaries and subsets. An
expert coder has many ways to successfully explore the data in an
aggregate lens using \texttt{agents\_to\_aggregate}. \pkg{EpiCompare}
also includes tools to rapidly explore data that has three epidemic
states. Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our
\texttt{geom\_aggregate} provides a rapid way to explore different
subpopulations' experience of an epidemic by combining the ideas behind
\texttt{agents\_to\_aggregate} for three-state models to examine
subpopulation trajectories in a 3d simplex space
\citep{Wickham2016, Hamilton2018}.
\textcolor{orange}{[Shannon says: come back?]}\footnote{\textcolor{violet}{[Ben says: what does this note mean?]}}
Visualization tools for three-state models were developed because (1)
SIR models are some of the most common and basic epidemic state-based
models and (2) our simplex representation of these epidemics emphasizes
a ``time-invarance'' representation of the data (for a refresher see
Section \ref{sec:time-invariant}).
\textcolor{orange}{[Shannon says: make sure SIR is defined before.]}
